The missing puzzle piece: contextual insights for enhanced pharmaceutical supply chain forecasting

### 2025-08-04 manuscript/03-experiment.qmd
3RD REVISION

For this study, we use a dataset spanning five years (Dec. 2017- July 2022) of consumption data obtained from EPSS. Rigorous checks were conducted to ensure the consistency and completeness of the collected data. From the extensive pool of pharmaceutical products within EPSS, we selected a set of 33 key pharmaceuticals, including various programs and representing different classes of drugs.

### Explaoratory Data Analysis with consumption data

Given the number of products in this study, we created several data plot and computed features of the time series, including the strength of trend and seasonality to better understadn data. @fig-feature shows the strength of trend versus seasonality. Each point represents one time series with the strength of trend in x-axis and the strength of seasonality in y-axis. Both measures are on a scale of [0,1]. the strength of trend and seasonality were calculated using the "STL" (Seasonal and Trend decomposition using Loess) decomposition method, as described by @bandara2025mstl

```{r}
#| label: fig-feature
#| cache: true
#| out.width: "60%"
#| fig.align: center
#| fig-cap: "The strength of the trend and seasonality in the time series of pharmaceutical product consumption. The scatter plot shows 33 data points, with each point corresponding to a product."

@@

It is evident that some time series display strong trends and/or seasonality, while the majority exhibit low trends and minimal seasonality. A number of products show pronounced trends, and only a few demonstrate clear seasonal patterns. Beyond assessing the strength of trends and seasonality, we also visualized all time series to understand data and various patterns, including trends and instances of erratic consumption behavior during certain months. For example, some series show low consumption volumes over consecutive months, followed by peak consumption in specific months, making them more volatile and difficult to forecast. This underscores the diversity of monthly pharmaceutical time series patterns within the dataset and highlights the importance of understanding the factors driving these consumption behaviors. Figure @fig-dataviz2 illustrates the time plots for a few selected products.

```{r}
#| label: fig-dataviz2
#| cache: true
#| dependson: "fig-feature"
#| fig-width: 11
#| fig-height: 6
#| fig-cap: "Monthly time plot of consumption. X-axis shows the month, consisting of 60 data points (months) and y-axis shows the comsumption. The panels display data from four products to give a glimpse of the consumption patterns."

p1 <- ggplot(med_qty |> filter(item == "Oxytocin") |> select(-item), 
       aes(x=month, y=quantity))+
  geom_point()+
  geom_line(aes(group =1))+
  scale_x_yearmonth(date_breaks = "4 month")+
  labs(x = "Month", y = "Consumption", title = "Oxytocin")+
  ggthemes::theme_few()+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

p2 <- ggplot(med_qty |> filter(item == "Amoxicillin - 500mg - Capsule") |> select(-item), 
       aes(x=month, y=quantity))+
  geom_point()+
  geom_line(aes(group =1))+
  scale_x_yearmonth(date_breaks = "4 month")+
    labs(x = "Month", y = "Consumption",title = "Amoxicillin - 500mg - Capsule")+
  ggthemes::theme_few()+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

p3 <- ggplot(med_qty |> filter(item == "Ceftriaxone") |> select(-item), 
       aes(x=month, y=quantity))+
  geom_point()+
  geom_line(aes(group =1))+
  scale_x_yearmonth(date_breaks = "4 month")+
    labs(x = "Month", y = "Consumption",title = "Ceftriaxone")+
  ggthemes::theme_few()+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

p4 <- ggplot(med_qty |> filter(item == "Atenolol - 50mg - Tablet") |> select(-item), 
       aes(x=month, y=quantity))+
  geom_point()+
  geom_line(aes(group =1))+
  scale_x_yearmonth(date_breaks = "4 month")+
    labs(x = "Month", y = "Consumption", title = "Atenolol - 50mg - Tablet")+
  ggthemes::theme_few()+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

@@

Visualizing the data revealed that various events significantly impacted the consumption of different products, but these were not reflected in the available data. Expert insights were crucial for understanding the nature and timing of these events, filling gaps in the system, and incorporating qualitative factors that EPSS logistics system miss. Therefore, we conducted interviews with domain experts to collect information on external factors that influence consumption. These interviews allowed us to account for unusual patterns and customize our forecasting models to more accurately reflect the unique consumption behaviors of each product.

@@

@tbl-feature also provides the summary statistics computed for each pharmaceutical product time series include traditional distributional measures—mean, median, standard deviation, minimum, and maximum—alongside time series–specific and structural characteristics. These include the first-order autocorrelation (ACF at lag 1), the mean length of consecutive zero runs (zero_run_mean), and the squared coefficient of variation computed only on non-zero values (nonzero_squared_cv). Together, these metrics capture not only central tendency and dispersion, but also temporal dependence, sparsity patterns, and relative variability in periods with non-zero consumption.

## Collaborative expert review to identify factors affecting consumption

\textcolor{blue}{To incorporate expert knowledge into time series forecasting in a structured and replicable manner, we developed a multi-step process involving expert engagement, collaborative data review, and operational validation. In partnership with the Forecasting and Market Shaping Directorate at EPSS, we identified six experienced professionals from the Warehouse and Inventory Management Directorate with deep operational knowledge of pharmaceutical logistics.}

Figure \ref{fig:expert-pipeline} \textcolor{blue}{presents a visual overview of the end-to-end approach, highlighting each major stage of the process from expert engagement to model integration.}

@@

\textcolor{blue}{Through a facilitated half-day workshop, these experts reviewed five years of monthly consumption data for 33 pharmaceutical products. Anomalous patterns—such as consumption spikes, prolonged lows, and erratic fluctuations—were jointly identified and interpreted in light of operational events (e.g., emergency distributions, inventory cycles). These insights were cross-validated against bin cards and warehouse records, and internal transfers were excluded to isolate externally relevant events. Dates were standardized to the Gregorian calendar.}

\textcolor{blue}{Predictor variables were defined through expert consensus, with inclusion contingent on agreement among all six core participants regarding the operational relevance and causal validity of each factor. The final predictors included binary indicators for stock replenishment events, categorical variables for fiscal year inventory counts, and dummy variables for malaria seasonality. A further description of these variables are summarized as followings:}

- \textcolor{blue}{Stock replenishment:  refers to the process of restocking or refilling inventory to ensure that there are sufficient quantities of products or materials available to meet demand. Whenever there was stock replenishment at the central EPSS, consumption and distribution to hubs and health facilities increased. This increase was attributed to the need to restock depleted inventories and the push from central EPSS to manage space constraints.}

- \textcolor{blue}{Physical fiscal year inventory counting: refers to the process of manually counting and verifying the actual quantities of pharmaceutical products available in stock at a specific location. The process is critical for maintaining the accuracy of inventory records, ensuring that medicines are available when needed, and preventing stockouts or overstocking.  Physical inventory counting periods also influenced consumption. Stores closed during these periods, halting transactions. We observed increased consumption before inventory counting periods, as hubs and facilities stocked up. July and August were identified as physical counting periods each year.}

- \textcolor{blue}{Malaria seasonality: Refers to the predictable patterns and fluctuations in malaria incidence throughout the year, typically influenced by climate and environmental conditions. In many regions, malaria transmission peaks during and shortly after the rainy season, when conditions such as stagnant water pools create ideal breeding sites for the Anopheles mosquitoes that transmit the disease. Conversely, malaria cases often decline during the dry season when mosquito breeding sites are reduced. During peak malaria seasons, there is a significant surge in the demand for antimalarial drugs and other related treatments. Malaria seasonality was another significant predictor. Certain pharmaceuticals, like Artemether + Lumefanthrine and Rapid Diagnostic Test kits, were affected by malaria outbreaks. We identified epidemic periods affecting consumption: September to December 2017, March to May 2018, September to December 2018, March to May 2019, September to December 2019, March to May 2020, September to December 2020, March to May 2021, September to December 2021, and March to May 2022.}

\textcolor{blue}{These predictors, encoded with known historical and future values, were added to the modeling dataset. This approach ensured consistent integration of expert-informed contextual variables across both point and probabilistic forecasting models.}

@@

**Exponential Smoothing State Space model (ETS):** ETS models, as described by @hyndman2021forecasting, combine trend, seasonality, and error components in time series using different configurations that can be additive, multiplicative, or mixed. The trend component can be specified as none ("N"), additive ("A"), or damped additive ("Ad"); the seasonality can be none ("N"), additive ("A"), or multiplicative ("M"); and the error term can be additive ("A") or multiplicative ("M"). To forecast consumption, we use the ETS() function from the fable package in R, which automatically selects the optimal model for each time series based on the corrected Akaike’s Information Criterion (AICc). In our study, an automated algorithm determines the best configuration for trend, seasonality, and error components across each time series, leveraging the ets() function’s use of AIC to identify the optimal model. Given the high volume of time series (1530), manual selection of components is impractical, so the algorithm customizes model forms for each series based on its unique characteristics. This results in a tailored combination of additive or multiplicative components, adapting to the specific patterns of each time series.

**Multiple Linear Regression (MLR)**: We use Multiple linear regression to model the relationship between a consumption and potential variables influencing its variation. In our first model, we use multiple linear regression with a trend component to capture the underlying progression over time, i.e., _regression_. We also incorporate dummy variables for each month to account for seasonal fluctuations, without including any additional predictors, i.e., _regression_reg_. This approach helps us establish a baseline model focused solely on temporal trends and seasonal patterns. We then extend this model by introducing additional predictors that include variables such as replenishment cycles, fiscal year indicators, and periods with malaria prevalence. These additional predictors allow the model to see if capturing external factors can provide a better understanding of the factors influencing the consumption and result at enhanced accuracy. We produce forecasts using TLSM() function from the fable package in R.

**ARIMA and ARIMA with regressors**: ARIMA (AutoRegressive Integrated Moving Average) is a powerful statistical model designed to forecast time series by capturing temporal dynamics and are widely used in time series forecasting due to their ability to model complex trends and patterns over time without relying on external predictors. ARIMAX (AutoRegressive Integrated Moving Average with eXogenous variables) extends ARIMA by incorporating external variables, or exogenous predictors, into the model. This modification allows to include relevant information from external factors such as malaria season, and fiscal year period, and stock replenishement period that may explain variations in the series beyond its internal time dynamics, we refer to this in the result as _arima_reg_. By adding these predictors, ARIMAX combines the strengths of ARIMA’s time-series structure with the flexibility of regression models. In our study, we use an automated algorithm to determine the optimal configuration for ARIMA components, following the approach outlined by @hyndman2021forecasting and We use the ARIMA() function from the fable package in R.

@@

\textcolor{blue}{To improve the robustness of the LSTM models and address overfitting issues, we introduced dropout regularization layers after the LSTM units and employed early stopping based on validation loss during model training. Each LSTM model was trained independently for each product series. Hyperparameter tuning was performed in a preliminary phase using a subset of products. Various configurations of LSTM units (30–100), dense units (50–200), dropout rates (0.1–0.5), and batch sizes (16–64) were evaluated based on validation set performance. The final model structure — 50 LSTM units, 100 dense units, a 0.2 dropout rate, and a batch size of 32 — was selected as a trade-off between forecast accuracy and model stability across different demand patterns.}

@@

\textcolor{blue}{In addition to point forecasts, we generated probabilistic forecasts to capture the uncertainty surrounding future pharmaceutical consumption. Several approaches are available for generating probabilistic forecasts, including analytical prediction intervals, quantile regression, Bayesian modeling through Markov Chain Monte Carlo (MCMC) methods, bootstrapping, and conformal prediction} [@wang2023].

\textcolor{blue}{In this study, we employed a bootstrapping approach to construct predictive intervals. Bootstrapping was chosen primarily for its flexibility and model-agnostic nature, allowing it to be applied uniformly across the diverse range of forecasting methods implemented without requiring strong distributional assumptions. Moreover, pharmaceutical consumption data often exhibit irregular and volatile patterns, making non-parametric approaches like bootstrapping particularly suitable.}

\textcolor{blue}{Specifically, we assume that future forecast errors will be similar to past forecast errors. The forecast error at time $t$ is defined as: $e_t = y_t - \hat{y}_t$, where $y_t$ represents the observed consumption, and $\hat{y}_t$ denotes the corresponding forecast estimate. To simulate future consumption paths, we randomly sample errors with replacement from the historical error distribution and add them to the point forecast estimates. This process is repeated multiple times (1,000 iterations in our study) to generate a distribution of possible outcomes for each forecast horizon.}

\textcolor{blue}{Prediction intervals at the desired confidence level (e.g., 95\%) are then constructed by taking appropriate quantiles from the empirical distribution of the simulated forecasts} [@hyndman2021forecasting].

\textcolor{blue}{The bootstrapping method thus provides a robust and flexible framework to quantify forecast uncertainty across a heterogeneous set of pharmaceutical products without imposing restrictive parametric assumptions.}

@@

\textcolor{blue}{To evaluate the performance of our forecasting models, we split the data into a series of training and test sets and employed a time series cross-validation approach, following best practices for forecasting evaluation} [@hyndman2021forecasting]. \textcolor{blue}{Rather than using a fixed training and test split, we used a rolling-origin cross-validation framework, which allows for a more comprehensive assessment across different demand patterns and periods. In our setup, the initial training set consisted of all available historical data from December 2017 up to June 2021. The evaluation period covered the subsequent 12 months, reflecting the operational needs of the EPSS, which plans consumption over a one-year horizon. At each iteration, models were trained on an expanding training window and evaluated over a fixed 6-month forecast horizon, aligned with typical EPSS planning cycles. After each forecast generation, the training set was expanded by one additional month, and the process was repeated, allowing for rolling assessment across the final 12 months of data. This structure ensured that forecasts reflected realistic operational scenarios, where forecasts are continuously updated as new data becomes available. This cross-validation design allowed us to evaluate each model's ability to perform across multiple different forecast origins and a variety of demand conditions, providing a more reliable and generalizable understanding of model performance. Furthermore, all model development steps, including hyperparameter tuning for the LSTM models, were conducted exclusively using the training data available at each iteration. No test set information was used during model selection or tuning.}

@@

To measure the forecast distribution accuracy, we calculate the Continuous Rank Probability Score [@hyndman2021forecasting]. It rewards sharpness and penalizes miscalibration, so it measures overall performance of the forecast distribution. For probabilistic evaluation, 1,000 future paths were simulated per series, enabling robust estimation of forecast uncertainty.

>>>>

For this study, we use a dataset spanning five years (December 2017 - July 2022) of distribution data obtained from EPSS. Rigorous checks were conducted to ensure the consistency and completeness of the collected data. From the extensive pool of pharmaceutical products within EPSS, we selected a set of 33 key pharmaceuticals, including various programs and representing different classes of drugs.

### Exploratory Data Analysis with distribution data

Given the number of products in this study, we created several data plot and computed features of the time series, including the strength of trend and seasonality to better understand data. @fig-feature shows the strength of trend versus seasonality. Each point represents one time series with the strength of trend in x-axis and the strength of seasonality in y-axis. Both measures are on a scale of [0,1]. the strength of trend and seasonality were calculated using the "STL" (Seasonal and Trend decomposition using Loess) decomposition method, as described by @bandara2025mstl

```{r}
#| label: fig-feature
#| cache: true
#| out.width: "60%"
#| fig.align: center
#| fig-cap: "The strength of the trend and seasonality in the time series of pharmaceutical product distribution. The scatter plot shows 33 data points, with each point corresponding to a product."

@@

It is evident that some time series display strong trends and/or seasonality, while the majority exhibit low trends and minimal seasonality. A number of products show pronounced trends, and only a few demonstrate clear seasonal patterns. Beyond assessing the strength of trends and seasonality, we also visualized all time series to understand data and various patterns, including trends and instances of erratic distribution behavior during certain months. For example, some series show low distribution volumes over consecutive months, followed by peak distribution in specific months, making them more volatile and difficult to forecast. This underscores the diversity of monthly pharmaceutical time series patterns within the dataset and highlights the importance of understanding the factors driving these distribution behaviors. @fig-dataviz2 illustrates the time plots for a few selected products.

```{r}
#| label: fig-dataviz2
#| cache: true
#| dependson: "fig-feature"
#| fig-width: 11
#| fig-height: 6
#| fig-cap: "Monthly time plot of distribution. X-axis shows the month, consisting of 60 data points (months) and y-axis shows the comsumption. The panels display data from four products to give a glimpse of the distribution patterns."

p1 <- ggplot(med_qty |> filter(item == "Oxytocin") |> select(-item), 
       aes(x=month, y=quantity))+
  geom_point()+
  geom_line(aes(group =1))+
  scale_x_yearmonth(date_breaks = "4 month")+
  labs(x = "Month", y = "Distribution", title = "Oxytocin")+
  ggthemes::theme_few()+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

p2 <- ggplot(med_qty |> filter(item == "Amoxicillin - 500mg - Capsule") |> select(-item), 
       aes(x=month, y=quantity))+
  geom_point()+
  geom_line(aes(group =1))+
  scale_x_yearmonth(date_breaks = "4 month")+
    labs(x = "Month", y = "Distribution",title = "Amoxicillin - 500mg - Capsule")+
  ggthemes::theme_few()+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

p3 <- ggplot(med_qty |> filter(item == "Ceftriaxone") |> select(-item), 
       aes(x=month, y=quantity))+
  geom_point()+
  geom_line(aes(group =1))+
  scale_x_yearmonth(date_breaks = "4 month")+
    labs(x = "Month", y = "Distribution",title = "Ceftriaxone")+
  ggthemes::theme_few()+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

p4 <- ggplot(med_qty |> filter(item == "Atenolol - 50mg - Tablet") |> select(-item), 
       aes(x=month, y=quantity))+
  geom_point()+
  geom_line(aes(group =1))+
  scale_x_yearmonth(date_breaks = "4 month")+
    labs(x = "Month", y = "Distribution", title = "Atenolol - 50mg - Tablet")+
  ggthemes::theme_few()+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

@@

Visualizing the data revealed that various events significantly impacted the distribution of different products, but these were not reflected in the available data. Expert insights were crucial for understanding the nature and timing of these events, filling gaps in the system, and incorporating qualitative factors that EPSS logistics system miss. Therefore, we conducted interviews with domain experts to collect information on external factors that influence distribution. These interviews allowed us to account for unusual patterns and customize our forecasting models to more accurately reflect the unique distribution behaviors of each product.

@@

@tbl-feature also provides the summary statistics computed for each pharmaceutical product time series include traditional distributional measures—mean, median, standard deviation, minimum, and maximum—alongside time series–specific and structural characteristics. These include the first-order autocorrelation (ACF at lag 1), the mean length of consecutive zero runs (zero_run_mean), and the squared coefficient of variation computed only on non-zero values (nonzero_squared_cv). Together, these metrics capture not only central tendency and dispersion, but also temporal dependence, sparsity patterns, and relative variability in periods with non-zero distribution.

## Collaborative expert review to identify factors affecting distribution

To incorporate expert knowledge into time series forecasting in a structured and replicable manner, we developed a multi-step process involving expert engagement, collaborative data review, and operational validation. In partnership with the Forecasting and Market Shaping Directorate at EPSS, we identified six experienced professionals from the Warehouse and Inventory Management Directorate with deep operational knowledge of pharmaceutical logistics.

Figure \ref{fig:expert-pipeline} presents a visual overview of the end-to-end approach, highlighting each major stage of the process from expert engagement to model integration.

@@

Through a facilitated half-day workshop, these experts reviewed five years of monthly distribution data for 33 pharmaceutical products. Anomalous patterns—such as distribution spikes, prolonged lows, and erratic fluctuations—were jointly identified and interpreted in light of operational events (e.g., emergency distributions, inventory cycles). These insights were cross-validated against bin cards and warehouse records, and internal transfers were excluded to isolate externally relevant events. Dates were standardized to the Gregorian calendar.

Predictor variables were defined through expert consensus, with inclusion contingent on agreement among all six core participants regarding the operational relevance and causal validity of each factor. The final predictors included binary indicators for stock replenishment events, categorical variables for fiscal year inventory counts, and dummy variables for malaria seasonality. A further description of these variables is summarized as followings:

- Stock replenishment:  refers to the process of restocking or refilling inventory to ensure that there are sufficient quantities of products or materials available to meet demand. Whenever there was stock replenishment at the central EPSS, distribution to hubs and health facilities increased. This increase was attributed to the need to restock depleted inventories and the push from central EPSS to manage space constraints.

- Physical fiscal year inventory counting: refers to the process of manually counting and verifying the actual quantities of pharmaceutical products available in stock at a specific location. The process is critical for maintaining the accuracy of inventory records, ensuring that medicines are available when needed, and preventing stockouts or overstocking.  Physical inventory counting periods also influenced distribution. Stores closed during these periods, halting transactions. We observed increased distribution before inventory counting periods, as hubs and facilities stocked up. July and August were identified as physical counting periods each year.

- Malaria seasonality: refers to the predictable patterns and fluctuations in malaria incidence throughout the year, typically influenced by climate and environmental conditions. In many regions, malaria transmission peaks during and shortly after the rainy season, when conditions such as stagnant water pools create ideal breeding sites for the Anopheles mosquitoes that transmit the disease. Conversely, malaria cases often decline during the dry season when mosquito breeding sites are reduced. During peak malaria seasons, there is a significant surge in the demand for antimalarial drugs and other related treatments. Malaria seasonality was another significant predictor. Certain pharmaceuticals, like Artemether + Lumefanthrine and Rapid Diagnostic Test kits, were affected by malaria outbreaks. We identified two recurrent epidemic seasons each year—March to May and September to December—that affected distribution from 2017 through 2022.

These predictors, encoded with known historical and future values, were added to the modeling dataset. This approach ensured consistent integration of expert-informed contextual variables across both point and probabilistic forecasting models.

@@

**Exponential Smoothing State Space model (ETS):** ETS models, as described by @hyndman2021forecasting, combine trend, seasonality, and error components in time series using different configurations that can be additive, multiplicative, or mixed. The trend component can be specified as none ("N"), additive ("A"), or damped additive ("Ad"); the seasonality can be none ("N"), additive ("A"), or multiplicative ("M"); and the error term can be additive ("A") or multiplicative ("M"). To forecast distribution, we use the ETS() function from the fable package in R, which automatically selects the optimal model for each time series based on the corrected Akaike’s Information Criterion (AICc). In our study, an automated algorithm determines the best configuration for trend, seasonality, and error components across each time series, leveraging the ets() function’s use of AIC to identify the optimal model. Given the high volume of time series (1530), manual selection of components is impractical, so the algorithm customizes model forms for each series based on its unique characteristics. This results in a tailored combination of additive or multiplicative components, adapting to the specific patterns of each time series.

**Multiple Linear Regression (MLR)**: We use Multiple linear regression to model the relationship between a distribution and potential variables influencing its variation. In our first model, we use multiple linear regression with a trend component to capture the underlying progression over time, i.e., _regression_. We also incorporate dummy variables for each month to account for seasonal fluctuations, without including any additional predictors, i.e., _regression_reg_. This approach helps us establish a baseline model focused solely on temporal trends and seasonal patterns. We then extend this model by introducing additional predictors that include variables such as replenishment cycles, fiscal year indicators, and periods with malaria prevalence. These additional predictors allow the model to see if capturing external factors can provide a better understanding of the factors influencing the distribution and result at enhanced accuracy. We produce forecasts using TLSM() function from the fable package in R.

**ARIMA and ARIMA with regressors**: ARIMA (AutoRegressive Integrated Moving Average) is a powerful statistical model designed to forecast time series by capturing temporal dynamics and are widely used in time series forecasting due to their ability to model complex trends and patterns over time without relying on external predictors. ARIMAX (Auto Regressive Integrated Moving Average with eXogenous variables) extends ARIMA by incorporating external variables, or exogenous predictors, into the model. This modification allows to include relevant information from external factors such as malaria season, and fiscal year period, and stock replenishement period that may explain variations in the series beyond its internal time dynamics, we refer to this in the result as _arima_reg_. By adding these predictors, ARIMAX combines the strengths of ARIMA’s time-series structure with the flexibility of regression models. In our study, we use an automated algorithm to determine the optimal configuration for ARIMA components, following the approach outlined by @hyndman2021forecasting and We use the ARIMA() function from the fable package in R.

@@

To improve the robustness of the LSTM models and address overfitting issues, we introduced dropout regularization layers after the LSTM units and employed early stopping based on validation loss during model training. Each LSTM model was trained independently for each product series. Hyperparameter tuning was performed in a preliminary phase using a subset of products. Various configurations of LSTM units (30–100), dense units (50–200), dropout rates (0.1–0.5), and batch sizes (16–64) were evaluated based on validation set performance. The final model structure — 50 LSTM units, 100 dense units, a 0.2 dropout rate, and a batch size of 32 — was selected as a trade-off between forecast accuracy and model stability across different demand patterns.

@@

In addition to point forecasts, we generated probabilistic forecasts to capture the uncertainty surrounding future pharmaceutical distribution Several approaches are available for generating probabilistic forecasts, including analytical prediction intervals, quantile regression, Bayesian modeling through Markov Chain Monte Carlo (MCMC) methods, bootstrapping, and conformal prediction [@wang2023].

In this study, we employed a bootstrapping approach to construct predictive intervals. Bootstrapping was chosen primarily for its flexibility and model-agnostic nature, allowing it to be applied uniformly across the diverse range of forecasting methods implemented without requiring strong distributional assumptions. Moreover, pharmaceutical distribution data often exhibit irregular and volatile patterns, making non-parametric approaches like bootstrapping particularly suitable.

Specifically, we assume that future forecast errors will be similar to past forecast errors. The forecast error at time $t$ is defined as: $e_t = y_t - \hat{y}_t$, where $y_t$ represents the observed distribution, and $\hat{y}_t$ denotes the corresponding forecast estimate. To simulate future distribution paths, we randomly sample errors with replacement from the historical error distribution and add them to the point forecast estimates. This process is repeated multiple times (1,000 iterations in our study) to generate a distribution of possible outcomes for each forecast horizon.

Prediction intervals at the desired confidence level (e.g., 95\%) are then constructed by taking appropriate quantiles from the empirical distribution of the simulated forecasts [@hyndman2021forecasting].

The bootstrapping method thus provides a robust and flexible framework to quantify forecast uncertainty across a heterogeneous set of pharmaceutical products without imposing restrictive parametric assumptions.

@@

To evaluate the performance of our forecasting models, we split the data into a series of training and test sets and employed a time series cross-validation approach, following best practices for forecasting evaluation [@hyndman2021forecasting]. Rather than using a fixed training and test split, we used a rolling-origin cross-validation framework, which allows for a more comprehensive assessment across different demand patterns and periods. In our setup, the initial training set consisted of all available historical data from December 2017 up to June 2021. The evaluation period covered the subsequent 12 months, reflecting the operational needs of the EPSS, which plans distribution over a one-year horizon. At each iteration, models were trained on an expanding training window and evaluated over a fixed 6-month forecast horizon, aligned with typical EPSS planning cycles. After each forecast generation, the training set was expanded by one additional month, and the process was repeated, allowing for rolling assessment across the final 12 months of data. This structure ensured that forecasts reflected realistic operational scenarios, where forecasts are continuously updated as new data becomes available. This cross-validation design allowed us to evaluate each model's ability to perform across multiple different forecast origins and a variety of demand conditions, providing a more reliable and generalizable understanding of model performance. Furthermore, all model development steps, including hyperparameter tuning for the LSTM models, were conducted exclusively using the training data available at each iteration. No test set information was used during model selection or tuning.

@@

To measure the forecast distribution accuracy, we calculate the Continuous Rank Probability Score (CRPS) [@hyndman2021forecasting]. It rewards sharpness and penalizes miscalibration, so it measures overall performance of the forecast distribution. For probabilistic evaluation, 1,000 future paths were simulated per series, enabling robust estimation of forecast uncertainty.
